Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms

### 2013-11-15 CGP Journal/response.txt

In [8], the authors use AND, OR, NAND, NOR for all of the other boolean problems tested (Even Parity, Digital Adder, and Digital Comparitor),
but use AND, AND with one input inverted, XOR and OR on digital multiplier.
The two papers [8] cites that performed research on this problem used AND, XOR.  They claim this makes the problem easier as
human designed solutions use only AND, XOR.  Therefore eliminating XOR as a choice likely makes this problem even harder.

It is possible to achieve O(1) amortized cost when performing mutations using the probability method.
This can be done leveraging the binomial distribution, which in many languages is implemented to give
the random number of trials k of N total which are true given probability P.  k genes can then
be selected from the genome at random just as when using a mutation rate.  This operator is
advantageous in that allows for a flexible number of mutating genes, up to and including
mutating all genes in a single step.  If I were to extend Single to use a configurable
number of genes, this would be my choice of method.

Over four thousand runs would be required to properly tune and evaluate any additional
problem.
A major focus of our work (ordering effect on the location and number of active genes)
is grounded in theory which makes no assumptions on the type of problem being solved.
As the experimental results matched the theory, we have no reason to suspect our
conclusions on these topics to change when the operators or data type is changed.
Some conclusions that could change depending on data type: The relative effectiveness
of the proposed modifications (IE, skip vs single, reorder vs normal, etc), the
evolved genome sizes, and the charts in Table IV and Table V.

The reason we chose all of our original problems as binary was to allow for the
detailed analysis of behavior in the latter half of Section VI.  If we had done
four problems, but using diverse domains, it would have been much more challenging
to determine why we observed changes in behavior between problems.  By using
four problems with similar domains and the same operators, we are able to determine
how our techniques scale with problem difficulty.

While we agree it is desirable to have more problems, and more diverse problems,
we were limited by how deep we wanted to go in analyzing our set of problems.
As stated elsewhere, each problem required approximately four thousand runs to
determine optimal genome size and mutation rate.  Similarly we wanted to delve
deeply into behavior, which requires time to determine measures and discover patterns,
not all of which are going to be domain independent.  As CGP's foundation is the
binary domain, we felt this was the correct choice for such detail oriented work.

Upon the suggestion of Reviewer 2, we further examined the statistical measures
of evaluations to success which we report in Table II.  As it is not safe to assume the number
of evaluations to success is normally distributed (in fact, they are most likely not normal),
we chose to provide non-parametric (median) instead of parametric (mean) measures,
as was done in [8].  Upon our review, we decided Median Absolute Deviation was not
capturing useful information, and we now show bootstrap confidence intervals, which is a much
more statistically sound and interesting value.

As to the including the wall clock time for our experiments, we attempted to cover our
reasoning for not doing this in the 4rd to last paragraph in the Results section.
In order to provide a fair comparison of runtime, we would need to redo tuning
with a focus on minimizing wall clock time instead of number of evaluations, which
would require another four thousand runs to time just one problem.  An additional
problem in reporting this value is hardware diversity.  To complete our experiments
we utilized a computing cluster with diverse hardware quality.  To avoid this problem,
we would need to recreate all experiments on homogeneous hardware, which is intractable
for experiments of this size.

### 2013-11-15 CGP Journal/main.tex
Completed a first pass of addressing reviewer comments

\subsection{Reorder}
\label{sec:reorder}
The \emph{Reorder} method shuffles an existing genome's node ordering without
changing node behavior.  This is possible because for a given node, there can
be a large number of other nodes which have no required ordering,
as they neither take input (directly or indirectly) from the node nor provide their
output (directly or indirectly) to the node.  In general, this process works by assigning nodes new
locations in the genome at random once all of the nodes they take input from have been assigned
locations earlier in the genome.

@@

\begin{table*}
	\centering
	\caption{Results for algorithm configuration on each problem.  Highlighted row
	is the control configuration \emph{Skip-Normal}.  MES=Median Evaluations to Success; MAD=Median
	Absolute Deviation; Active=Median number of active nodes in evolved solution;
	Used=Median number of nodes active in evolved solution after applying SIMPLIFY (See Section~\ref{sec:behavior});
	p-value=Result from Mann-Whitney~U comparison with control.
	}
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
	  \hline
	  \multicolumn{9}{|c|}{\textbf{3-bit Parity}, Kruskal-Wallis=$5.99\cdot 10^{-14}$} \\ \hline
\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 348 & 122 & 20 & 0.5602 \\ \hline
\emph{Accumulate} & \emph{Reorder}&  5,000 & 0.01 & 796 & 434 & 331 & 23 & 0.7174 \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}       & 10,000 & 0.02 & 823 & 363 & 119 & 20 & NA \\ \hline
\emph{Skip} & \emph{Reorder}      &  2,000 & 0.01 & 870 & 362 & 198 & 20 & 0.3125 \\ \hline
\emph{Accumulate} & \emph{Dag}    &  5,000 & 0.01 & 964 & 499 & 1,174 & 22 & 0.3310 \\ \hline
\emph{Skip} & \emph{Dag}          & 10,000 & 0.02 & 1,042 & 620 & 2,083 & 24 & 0.0248 \\ \hline
\emph{Single} & \emph{Normal}     & 500 & NA & 1,214 & 497 & 39 & 17 & 0.0046 \\ \hline
\emph{Single} & \emph{Reorder}    & 200 & NA & 1,506 & 823 & 57 & 18 & 0 \\ \hline
\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 1,313 & 43 & 15 & 0 \\ \hline\hline

@@

\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}        & 10,000 & 0.005 & 16,226 & 7,455  &   283 & 49 & NA \\ \hline
\emph{Skip} & \emph{Reorder}       & 10,000 & 0.005 & 17,883 & 8,378  & 2,070 & 62 & 0.6918 \\ \hline
\emph{Accumulate} & \emph{Reorder} & 10,000 & 0.002 & 18,132 & 8,926  & 2,027 & 64 & 0.8822 \\ \hline
\emph{Accumulate} & \emph{Normal}  & 10,000 & 0.005 & 21,428 & 6,379  &   279 & 49 & 0.2525 \\ \hline
\emph{Accumulate} & \emph{Dag}     & 10,000 & 0.002 & 24,855 & 12,512 & 4,059 & 69 & 0.0736 \\ \hline
\emph{Skip} & \emph{Dag}           & 10,000 & 0.002 & 25,678 & 12,034 & 4,178 & 62 & 0.0211 \\ \hline
\emph{Single} & \emph{Reorder}     & 100    & NA    & 26,558 & 7,858  &    52 & 32 & 0.0076 \\ \hline
\emph{Single} & \emph{Normal}      & 2,000  & NA    & 31,057 & 14,475 &   141 & 43 & 0.0085 \\ \hline
\emph{Single} & \emph{Dag}         & 100    & NA    & 34,208 & 10,069 &    59 & 34 & 0 \\ \hline\hline

@@

\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\emph{Skip} & \emph{Reorder}       &  2,000 & 0.002 &  62,853 & 20,023 &   757 & 114 & 0.5247 \\ \hline
\emph{Single} & \emph{Reorder}     &    500 &    NA &  63,731 & 13,119 &   245 &  97 & 0.3610 \\ \hline
\emph{Accumulate} & \emph{Normal}  &  1,000 & 0.002 &  67,390 & 11,991 &   211 &  87 & 0.8767 \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}        &  1,000 & 0.002 &  68,231 & 15,115 &   202 &  85 & NA \\ \hline
\emph{Single} & \emph{Normal}      &  1,000 &    NA &  68,819 & 14,590 &   203 &  85 & 0.7695 \\ \hline
\emph{Accumulate} & \emph{Reorder} &  2,000 & 0.002 &  72,991 & 18,650 &   760 & 115 & 0.1868 \\ \hline
\emph{Skip} & \emph{Dag}           & 10,000 & 0.001 & 132,821 & 29,432 & 4,983 & 125 & 0 \\ \hline
\emph{Accumulate} & \emph{Dag}     &  5,000 & 0.002 & 133,188 & 29,326 & 2,542 & 125 & 0 \\ \hline
\emph{Single} & \emph{Dag}         &    500 &    NA & 160,696 & 39,242 &   299 &  97 & 0 \\ \hline\hline

@@

\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\emph{Skip} & \emph{Reorder}      & 10,000 & 0.0005  & 196,116  & 60,326  & 3,175 & 484 & 0  \\ \hline
\emph{Accumulate} & \emph{Reorder}& 10,000 & 0.001 & 199,193 & 85,970 & 3,117 & 489 & 0  \\ \hline
\emph{Single} & \emph{Dag} & 200  & NA & 239,844 & 78,739 & 114 & 74 & 0 \\ \hline
\emph{Single} & \emph{Reorder}    & 500   & NA & 243,664 & 91,935 & 266 & 131 & 0 \\ \hline
\emph{Skip} & \emph{Dag}          & 2,000 & 0.002 & 369,931 & 174,408 & 952 & 149 & 0.0100 \\ \hline
\emph{Accumulate} & \emph{Dag}    & 2,000 & 0.002  & 379,202 & 147,289 & 973 & 158 & 0.0053  \\ \hline
\emph{Single} & \emph{Normal}     & 5,000  & NA & 399,911 & 194,384 & 273 & 112 & 0.0047 \\ \hline
\emph{Accumulate} & \emph{Normal} & 2,000 & 0.002 & 483,938 & 227,686 & 168 & 87 & 0.45 \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal} & 2,000 & 0.002 & 657,121 & 311,376 & 175 & 84 & NA \\ \hline

@@

The results from running each tuned parameter configuration for each algorithm
combination on each problem are summarized by Table~\ref{tab:results}.
Rows for each problem are ordered by that algorithm combination's median evaluations to
success (MES), with the best algorithm combination appearing at the top of each section.
The highlighted rows mark \emph{Skip-Normal}, the control algorithm combination
representing traditional CGP performance.  All configurations for each problem
were then tested using Kruskal-Wallis to determine if any configuration was
statistically different from any other.  All problems were determined to be highly significant.
As such, each algorithm combination was then compared with the control
to determine statistical significance using pairwise Mann-Whitney~U tests. The
resulting p-value for each test is reported in the table.

@@

As a final check that our results do not somehow unfairly handicap traditional CGP,
we compared our median evaluations to success with previous publications.  The results
reported for traditional CGP and CGP using modules in~\cite{walker:2008:cgpmodules}
for 3-bit parity required 6 times as many evaluations as \emph{Skip-Normal}.
The same source had traditional CGP requiring 4 million evaluations to success,
as opposed to our \emph{Skip-Normal} requiring 600 thousand, also a factor of 6 improvement.
CGP with modules required over 800 thousand evaluations, worse than our \emph{Skip-Normal},
and approximately 4 times more than \emph{Skip-Reorder}.  We must be wary about comparisons
to their 3-bit multiplier though, as their version used a different function set.
That said the function set they used appears to have been chosen explicitly to improve
CGP's performance on that problem, so we suggest that if we had used a the same function
set our results would still be significantly better.

>>>>

%NEW
It is important to note that Equation~\ref{eq:prob} does not apply to all variants of CGP.
For instance, the early versions of CGP~\cite{miller:2011:chapter2} which utilize a
levels back parameter $L$ will have
a flat active probability for all nodes at least $L$ nodes from the start and $L$ nodes from the end
of the genome.  Yet this does not remove the issue of artificially limiting connections,
and in fact adds many new limitations.  More recently,~\cite{cai:2005:ircgp} proposed
a method which reassembled the genome using semantic cues, which circumvents the
forward only limitation in \emph{Normal}.
%/NEW
\subsection{Reorder}
\label{sec:reorder}
The \emph{Reorder} method shuffles an existing genome's node ordering without
changing node behavior.  This is possible because for a given node, there can
be a large number of other nodes which have no required ordering,
as they neither take input (directly or indirectly) from the node nor provide their
output (directly or indirectly) to the node.  In general, this process works by assigning nodes new
locations in the genome at random once all of the nodes they take input from have been assigned
locations earlier in the genome.

@@

\begin{table*}
	\centering
	\caption{Results for algorithm configuration on each problem.  Highlighted row
	is the control configuration \emph{Skip-Normal}.  MES=Median Evaluations to Success;
	Confidence Interval=95\% bootstrapped confidence interval; Active=Median number of active nodes in evolved solution;
	Used=Median number of nodes active in evolved solution after applying SIMPLIFY (See Section~\ref{sec:behavior});
	p-value=Result from Mann-Whitney~U comparison with control.
	}
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
	  \hline
	  \multicolumn{9}{|c|}{\textbf{3-bit Parity}, Kruskal-Wallis=$5.99\cdot 10^{-14}$} \\ \hline
\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{Confidence Interval} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 445 .. 869 & 122 & 20 & 0.5602 \\ \hline
\emph{Accumulate} & \emph{Reorder}&  5,000 & 0.01 & 796 & 576 .. 1,019 & 331 & 23 & 0.7174 \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}       & 10,000 & 0.02 & 823 & 662 .. 1,034 & 119 & 20 & NA \\ \hline
\emph{Skip} & \emph{Reorder}      &  2,000 & 0.01 & 870 & 708 .. 1,070 & 198 & 20 & 0.3125 \\ \hline
\emph{Accumulate} & \emph{Dag}    &  5,000 & 0.01 & 964 & 759 .. 1,250 & 1,174 & 22 & 0.3310 \\ \hline
\emph{Skip} & \emph{Dag}          & 10,000 & 0.02 & 1,042 & 734 .. 1,339 & 2,083 & 24 & 0.0248 \\ \hline
\emph{Single} & \emph{Normal}     & 500 & NA & 1,214 & 885 .. 1,487 & 39 & 17 & 0.0046 \\ \hline
\emph{Single} & \emph{Reorder}    & 200 & NA & 1,506 & 552 .. 2,122 & 57 & 18 & 0 \\ \hline
\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 2,109 .. 3,358 & 43 & 15 & 0 \\ \hline\hline

@@

\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{Confidence Interval} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}        & 10,000 & 0.005 & 16,226 & 9,315 .. 22,229  &   283 & 49 & NA \\ \hline
\emph{Skip} & \emph{Reorder}       & 10,000 & 0.005 & 17,883 & 10,690 .. 23,999 & 2,070 & 62 & 0.6918 \\ \hline
\emph{Accumulate} & \emph{Reorder} & 10,000 & 0.002 & 18,132 & 13,646 .. 22,983 & 2,027 & 64 & 0.8822 \\ \hline
\emph{Accumulate} & \emph{Normal}  & 10,000 & 0.005 & 21,428 & 17,536 .. 25,948 &   279 & 49 & 0.2525 \\ \hline
\emph{Accumulate} & \emph{Dag}     & 10,000 & 0.002 & 24,855 & 18,222 .. 33,069 & 4,059 & 69 & 0.0736 \\ \hline
\emph{Skip} & \emph{Dag}           & 10,000 & 0.002 & 25,678 & 16,170 .. 33,422 & 4,178 & 62 & 0.0211 \\ \hline
\emph{Single} & \emph{Reorder}     & 100    & NA    & 26,558 & 20,678 .. 31,613 &    52 & 32 & 0.0076 \\ \hline
\emph{Single} & \emph{Normal}      & 2,000  & NA    & 31,057 & 24,145 .. 42,278 &   141 & 43 & 0.0085 \\ \hline
\emph{Single} & \emph{Dag}         & 100    & NA    & 34,208 & 29,840 .. 38,035 &    59 & 34 & 0 \\ \hline\hline

@@

\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{Confidence Interval} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\emph{Skip} & \emph{Reorder}       &  2,000 & 0.002 &  62,853 & 50,033 .. 72,267 &   757 & 114 & 0.5247 \\ \hline
\emph{Single} & \emph{Reorder}     &    500 &    NA &  63,731 & 56,070 .. 70,706 &   245 &  97 & 0.3610 \\ \hline
\emph{Accumulate} & \emph{Normal}  &  1,000 & 0.002 &  67,390 & 60,738 .. 74,624 &   211 &  87 & 0.8767 \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}        &  1,000 & 0.002 &  68,231 & 63,266 .. 74,463 &   202 &  85 & NA \\ \hline
\emph{Single} & \emph{Normal}      &  1,000 &    NA &  68,819 & 62,940 .. 74,939 &   203 &  85 & 0.7695 \\ \hline
\emph{Accumulate} & \emph{Reorder} &  2,000 & 0.002 &  72,991 & 67,163 .. 78,131 &   760 & 115 & 0.1868 \\ \hline
\emph{Skip} & \emph{Dag}           & 10,000 & 0.001 & 132,821 & 117,034 .. 155,471 & 4,983 & 125 & 0 \\ \hline
\emph{Accumulate} & \emph{Dag}     &  5,000 & 0.002 & 133,188 & 121,369 .. 147,073 & 2,542 & 125 & 0 \\ \hline
\emph{Single} & \emph{Dag}         &    500 &    NA & 160,696 & 142,001 .. 160,696 &   299 &  97 & 0 \\ \hline\hline

@@

\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{Confidence Interval} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\emph{Skip} & \emph{Reorder}      & 10,000 & 0.0005& 196,116 & 160,073 .. 223,500  & 3,175 & 484 & 0  \\ \hline
\emph{Accumulate} & \emph{Reorder}& 10,000 & 0.001 & 199,193 & 123,496 .. 275,013 & 3,117 & 489 & 0  \\ \hline
\emph{Single} & \emph{Dag}        &   200  & NA    & 239,844 & 204,347 .. 286,418 & 114 & 74 & 0 \\ \hline
\emph{Single} & \emph{Reorder}    & 500    & NA    & 243,664 & 178,566 .. 301,246 & 266 & 131 & 0 \\ \hline
\emph{Skip} & \emph{Dag}          & 2,000  & 0.002 & 369,931 & 234,885 .. 483,753 & 952 & 149 & 0.0100 \\ \hline
\emph{Accumulate} & \emph{Dag}    & 2,000  & 0.002 & 379,202 & 255,165 .. 460,783 & 973 & 158 & 0.0053  \\ \hline
\emph{Single} & \emph{Normal}     & 5,000  & NA    & 399,911 & 280,070 .. 516,034 & 273 & 112 & 0.0047 \\ \hline
\emph{Accumulate} & \emph{Normal} & 2,000  & 0.002 & 483,938 & 308,049 .. 611,034 & 168 & 87 & 0.45 \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}       & 2,000  & 0.002 & 657,121 & 469,287 .. 864,073 & 175 & 84 & NA \\ \hline

@@

The results from running each tuned parameter configuration for each algorithm
combination on each problem are summarized by Table~\ref{tab:results}.
Rows for each problem are ordered by that algorithm combination's median evaluations to
success (MES), with the best algorithm combination appearing at the top of each section.
The highlighted rows mark \emph{Skip-Normal}, the control algorithm combination
representing traditional CGP performance.  All configurations for each problem
were then tested using
%NEW
Kruskal-Wallis\footnote{Kruskal-Wallis, a non-parametric equivalent of ANOVA, determines
if samples were drawn from the same distribution.  If this test is significant, at
least one of the samples differs from the others.}
%/NEW
to determine if any configuration was
statistically different from any other.  All problems were determined to be highly significant.
As such, each algorithm combination was then compared with the control
to determine statistical significance using pairwise
%NEW
Mann-Whitney~U\footnote{Mann-Whitney~U, a non-parametric equivalent of Student's t-test,
determines if two populations were drawn from the same distribution.  If this test is significant,
the populations are not equal.}
%/NEW
tests. The
resulting p-value for each test is reported in the table.

@@

As a final check that our results do not somehow unfairly handicap traditional CGP,
we compared our median evaluations to success with previous publications.  The results
reported for traditional CGP in~\cite{walker:2008:cgpmodules}
for 3-bit parity required 6 times as many evaluations as \emph{Skip-Normal}.
On the 3-bit Multiplier problem using a slightly different operator set,
the same source had traditional CGP requiring 4 million evaluations to success,
as opposed to our \emph{Skip-Normal} requiring 600 thousand, also a factor of 6 improvement.
%NEW
To determine the effect the difference in function set had on results, we ran
\emph{Skip-Normal} and \emph{Skip-Reorder} using their functions but our parameters.
The former improved to 453,684 while the latter increased to 303,657, maintaining
a statistically significant difference.  These
results make sense as the operator set in~\cite{walker:2008:cgpmodules} is able
to more compactly represent the solution than our set.  This puts less strain
on \emph{Normal} to deal with complex solutions, while \emph{Reorder}'s extremely
large genome size is likely made unnecessary.  Note that as we did not tune
using their operator set, our results should be considered a lower bound on quality for
using \emph{Skip-Normal} and \emph{Skip-Reorder} to solve 3-bit Multiplier
with their operators.
%/NEW
